Dynamics of innovation in stigmergic collectives

### 2023-05-07 LaTeX/draft.tex
HomeEdits

\subsection*{Search strategy post bifurcation}
After the agent leaves the trail trajectory at $\r(s_c)$, the agent needs to reach the target A.
Though the agent can take a variety of exploration strategies, we assume that the policy resembles
an Ornstein-Uhlenbeck (OU) process in the agent's orientation. This fits well with the behavior of an insect such as an ant because
ants often explore using their orientaiton after making measurements with their antennae.
This explorative policy $\pi_{\OU}(\r)$ can be written through $\P(\theta, t)$
the probability that the agent will take a particular angle $\theta$ at time $t$ or equivalently at distance $s$.
However, since the orientation of the agent is dependent on $\r(s_c)$, the angle between agent location $\r(s_c)$
to target A can be denoted as $\phi$ and $\theta=\phi+\varphi$ where $\varphi$ is the fluctuation around this
orientation that follows an OU-process. The probability $\P(\varphi, s)$ can then be written as
\[
    \frac{\partial \P}{\partial s} = \frac{\nu}{v_o} \frac{\partial(\varphi \P)}{\partial \varphi }
    + \frac{D}{v_o} \frac{\partial^2 \P}{\partial \varphi^2}
\]
where $\nu$ is the time-scale associated with relaxation of the agent's orientation to the target $\phi$
and $D$ represents the fluctuation in orientation, which we will soon assume to be large when $\r(s_c)$
is farther from the target A. Often in navigating insects, agents can quickly detect food source when they
are close to the food source. We capture this by assuming that after reaching a distance $\sigma$ from the end-point A,
agents can reach the target and thus the reward is a function only of the variance in agent's position towards the
target. Instead of solving the partial differential equation for $\P(\varphi, s)$, we can evaluate the
evolution of the variance of agent's position $\r(s)$ as it marches close to the target. Let $\ebh_\phi$
denote the orientation of the agent from $\r(s_c)$ to the target A. We can write the evolution of the variance
along $\ebh_\perp$ i.e. $\psi(s)=\E[r^2_\perp(s)]$ as $\d{\psi}(s)/\d s = 2 \E [ r_\perp(s) \varphi ]$ for $s>s_c$.
We can evaluate this evolution to get
\begin{align}
    \psi(s) - \psi(s_c) & = \frac{v_o^2 D}{\nu^3} \bigg[ \frac{2 \nu (s-s_c)}{v_o} - \e^{-2 \nu (s-s_c)/v_o} \nonumber \\
    & \ + 4 \e^{-\nu (s-s_c)/v_o} - 3 \bigg].
\end{align}
Before we move further with the calculation, it is cogent to work in non-dimensional numbers. In the OU-process
we have two length-scales: $l \sim (v_o/\nu), (v_o/D)$, one due to \textit{inertia} and other due to \textit{diffusion}.
The ratio of these two is nothing but the Peclet number, $\Pe = (D/\nu)$. We choose $l_\nu = (v_o/\nu)$ as the length-scale
for further non-dimensionalization such that $s \rightarrow l_\nu \hat{s}$. We drop the hat in what follows and we get
the non-dimensional variance, $\psi(s)$ (after setting $\psi(s_c) =0$) as a function of non-dimensional distance, $s$ as
\begin{align}
    \psi(s)& = \Pe \bigg[ 2 (s-s_c) - \e^{-2 (s-s_c)} + 4 \e^{-(s-s_c)} - 3 \bigg].
\end{align}
As already mentioned, the agent reaching a distance $\sigma$ (appropriately non-dimensionalized with $l_\nu$) along $\ebh_\perp$ from the end-point A
can reach the target, the probability that an agent will reach the target is given by $\P(||r_\perp^{s^*} - r^*|| \leq \sigma)
= \sigma/\psi(s^*)$. We can now calculate the value function, $v_{\rm OU}(\r, \theta)$ as
\[
    v_{\rm OU}(\r) = \int_{s_c}^{s^*} \int \D[\theta(s)] \P_{\rm OU}[\theta(s)] R(s) \ \d s.
\]
The reward function for the OU process takes a simple form $R(s) = \mu[1 - \Theta(||r_\perp(s^*) - r^*|| - \sigma)]$,
which simply means that the agent reaching with a distance $\sigma$ gets a reward $\mu$.
From this we see that the value of $v_{\rm OU}(\r^c) = \sigma \mu/\psi(s^*)$.

\section*{Value function of combined policy}
The total value function for the agent's policy is $v(\r, \theta) = v_{\rm int}(\r, \theta) + v_{\rm OU}(\r^c, \theta^c)$.
The agent is trying to maximize this value function from a given state, $\r^t, \theta^t$. The final expression
for the value function is
\[
    v(\r; s_c, D) = \frac{\sigma \mu}{\psi(s^*; s_c, \Pe)} + \frac{\phi^*}{\kappa} [ \e^{-\kappa s/\phi^*} - \e^{-\kappa s_c/\phi^*}],
\]
where we have introduced the non-dimensional number, $\kappa = \kappa_o l_\nu = (\kappa_o v_o/\nu)$.
In this equation, the Peclet number $\Pe$ captures the fluctuations in the dynamics of the agent
associated with the error in the homing vector. We assume that this diffusion is larger farther
from the target location and takes the functional form $\Pe \sim 1/s$. Moreover the reward that the agent recieves
when the bifurcation point is close to the target location needs to be smaller and thus $\mu = (\pi\kappa^{-1} -s_c)$.
One last detail that needs to be specified to make the value function complete is to specify $s^*$. It is straight
forward to show that from $s_c$, the target A is a distance $2 \cos(s_c\kappa/2)\kappa^{-1}$ which makes
$s^* = s_c + 2 \cos(s_c\kappa/2)\kappa^{-1}$.

>>>>

\subsection*{Search strategy post bifurcation}
After the agent leaves the trail trajectory at $\r(s_c)$, the agent needs to reach the target A.
Though the agent can take a variety of exploration strategies, we assume that the policy resembles
an Ornstein-Uhlenbeck (OU) process in the agent's orientation. This fits well with the behavior of an insect such as an ant because
ants often explore using their orientaiton after making measurements with their antennae.
This explorative policy $\pi_{\OU}(\r)$ can be written through $\P(\theta, t)$
the probability that the agent will take a particular angle $\theta$ at time $t$ or equivalently at distance $s$.
However, since the orientation of the agent is dependent on $\r(s_c)$, the angle between agent location $\r(s_c)$
to target A can be denoted as $\phi$ and $\theta=\phi+\varphi$ where $\varphi$ is the fluctuation around this
orientation that follows an OU-process. The probability $\P(\varphi, s)$ can then be written as
\[
    \frac{\partial \P}{\partial s} = \frac{\nu}{v_o} \frac{\partial(\varphi \P)}{\partial \varphi }
    + \frac{D}{v_o} \frac{\partial^2 \P}{\partial \varphi^2}
\]
where $\nu$ is the time-scale associated with relaxation of the agent's orientation to the target $\phi$
and $D$ represents the fluctuation in orientation, which we will soon assume to be large when $\r(s_c)$
is farther from the target A. Often in navigating insects, agents can quickly detect food source when they
are close to the food source. We capture this by assuming that after reaching a distance $\sigma$ from the end-point A,
agents can reach the target and thus the reward is a function only of the variance in agent's position towards the
target. Instead of solving the partial differential equation for $\P(\varphi, s)$, we can evaluate the
evolution of the variance of agent's position $\r(s)$ as it marches close to the target. Let $\ebh_\phi$
denote the orientation of the agent from $\r(s_c)$ to the target A. We can write the evolution of the variance
along $\ebh_\perp$ i.e. $\psi(s)=\E[r^2_\perp(s)]$ as $\d{\psi}(s)/\d s = 2 \E [ r_\perp(s) \varphi ]$ for $s>s_c$.
We can evaluate this evolution to get
\begin{align}
    \psi(s) - \psi(s_c) & = \frac{v_o^2 D}{\nu^3} \bigg[ \frac{2 \nu (s-s_c)}{v_o} - \e^{-2 \nu (s-s_c)/v_o} \nonumber \\
    & \ + 4 \e^{-\nu (s-s_c)/v_o} - 3 \bigg].
\end{align}
Before we move further with the calculation, it is cogent to work in non-dimensional numbers. In the OU-process
we have two length-scales: $l \sim (v_o/\nu), (v_o/D)$, one due to \textit{inertia} and other due to \textit{diffusion}.
The ratio of these two is nothing but the Peclet number, $\Pe = (D/\nu)$. We choose $l_\nu = (v_o/\nu)$ as the length-scale
for further non-dimensionalization such that $s \rightarrow l_\nu \hat{s}, t \rightarrow \nu^{-1} \hat{t}$. We drop the hat in what follows and we get
the non-dimensional variance, $\psi(s)$ (after setting $\psi(s_c) =0$) as a function of non-dimensional distance $s$ to be
\begin{align}
    \psi(s)& = \Pe \bigg[ 2 (s-s_c) - \e^{-2 (s-s_c)} + 4 \e^{-(s-s_c)} - 3 \bigg].
\end{align}
As already mentioned, the agent reaching a distance $\sigma$ (appropriately non-dimensionalized with $l_\nu$) along $\ebh_\perp$ from the end-point A
can reach the target, then the probability that an agent will reach the target is given by $\P(||r_\perp^{s^*} - r^*|| \leq \sigma)
= \sigma/\psi(s^*)$. We can now calculate the value function, $v_{\rm OU}(\r, \theta)$ as
\[
    v_{\rm OU}(\r) = \int_{s_c}^{s^*} \int \D[\theta(s)] \P_{\rm OU}[\theta(s)] R(s) \ \d s.
\]
The reward function for the OU process takes a simple form $R(s) = \mu[1 - \Theta(||r_\perp(s^*) - r^*|| - \sigma)]$,
which simply means that the agent reaching within a distance $\sigma$ gets a reward $\mu$.
From this we see that the value function from $\r$ to the end of the trajectory, $v_{\rm OU}(\r^c) = \sigma \mu/\psi(s^*)$.

\section*{Value function of combined policy}
The total value function for the agent's policy is simply $v(\r, \theta) = v_{\rm int}(\r, \theta) + v_{\rm OU}(\r^c, \theta^c)$.
The agent is trying to maximize this value function from a given state, $\r^t, \theta^t$. The final expression
for the value function is
\[
    v(\r; s_c, D) = \frac{\sigma \mu}{\psi(s^*; s_c, \Pe)} + \frac{\phi^*}{\kappa} [ \e^{-\kappa s/\phi^*} - \e^{-\kappa s_c/\phi^*}],
\]
where we have introduced the non-dimensional number, $\kappa = \kappa_o l_\nu = (\kappa_o v_o/\nu)$.
In this equation, the Peclet number $\Pe$ captures the fluctuations in the dynamics of the agent
associated with the error in the homing vector. We assume that this diffusion is larger farther
from the target location and takes the functional form $\Pe \sim 1/s$. Moreover the reward that the agent recieves
when the bifurcation point is close to the target location needs to be smaller and thus $\mu = (\pi\kappa^{-1} -s_c)$.
One last detail that needs to be specified to make the value function complete is to specify $s^*$. It is straight
forward to show that from $s_c$, the target A is a distance $2 \cos(s_c\kappa/2)\kappa^{-1}$ which makes
$s^* = s_c + 2 \cos(s_c\kappa/2)\kappa^{-1}$.
